---
title: "GSOC Tests(2019)"
author: "Anuraag Srivastava(as4378@nau.edu)"
date: "February 22, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

**Test 1**


Loading the required packages:


```{r}
library(data.table)
library(dplyr)
library(changepoint)
library(ggplot2)
library(microbenchmark)
library(fpop)
data(neuroblastoma, package="neuroblastoma")
```
\vspace{1cm}

Selecting one profile id and one chromosome to continue with the test.


```{r}
selected_profile_id = "1"
selected_chromosome = "1"
```
\vspace{1cm}

Creating a data table for selected profile:


```{r}
selected.profiles <- data.table(filter(neuroblastoma$profiles, 
                                       profile.id == selected_profile_id, 
                                       chromosome == selected_chromosome))
```

\vspace{2cm}

Plotting this problem in a grid we get:


```{r}

gg.unsupervised <- ggplot()+
  theme(
    panel.margin=grid::unit(0, "lines"),
    panel.border=element_rect(fill=NA, color="grey50")
  )+
  facet_grid(profile.id ~ chromosome, scales="free", space="free_x")+
  geom_point(aes(position, logratio),
             data=selected.profiles,
             shape=1)+
  scale_x_continuous(
    "position on chromosome",
    breaks=c(100, 200))+
  scale_y_continuous(
    "logratio",
    limits=c(-1,1)*1.1)
print(gg.unsupervised)
```


Now, fitting the unsupervised change point model using the cpt.mean function and using "SIC0" as the penalty parameter we get:

\vspace{1cm}
```{r}
pen.name <- "SIC0"
(models <- selected.profiles[, {
  fit.pelt <- cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(as.integer((position[before.change]+position[after.change])/2)),
    before_mean=mean(logratio[1:before.change]),
    after_mean=mean(logratio[after.change:length(logratio)]),
    end_pos=position[length(position)]
    )
}, by=list(profile.id, chromosome)])


(changes <- models[, data.table(
  change=changes[[1]],before_mean = before_mean, after_mean = after_mean, end_pos=end_pos
), by=list(profile.id, chromosome, pen.name)])

segments.mean = NULL
segments.mean = rbind(segments.mean, c("start" = 0, "end" = changes$change,
                                       "mean" = changes$before_mean))
segments.mean = rbind(segments.mean, c("start" = changes$change, 
                                       "end" = changes$end_pos,
                                       "mean" = changes$after_mean))
```


\vspace{1cm}
Plotting the data along with optimal segment means (green line segments) we get:


```{r}
gg.unsupervised+
  theme(legend.box="horizontal")+
  geom_vline(aes(
    xintercept=change),
    color="green",
    size=1,
    linetype="dashed",
    data=changes)+
  geom_segment(aes(
    x = start, 
    y = mean,
    xend = end,
    yend = mean,
    col = I("green")), 
    data = as.data.frame(segments.mean))
```



**Test 2**


Seleceting 2 data sets.


```{r}
profile1 = "1"
profile2 = "4"
chromosome1 = "1"
chromosome2 = "1"
```

\vspace{1cm}
Creating data table for selected profiles:


```{r}
selected.profiles1 = filter(neuroblastoma$profiles, profile.id == profile1, 
                             chromosome == chromosome1)
selected.profiles2 = filter(neuroblastoma$profiles, profile.id == profile2, 
                             chromosome == chromosome2)

selected.profiles = data.table(rbind(selected.profiles1, selected.profiles2))
```

\vspace{1cm}
Printing the problems:

```{r}
gg.unsupervised <- ggplot()+
  theme(
    panel.margin=grid::unit(0, "lines"),
    panel.border=element_rect(fill=NA, color="grey50")
  )+
  facet_grid(profile.id ~ chromosome, scales="free", space="free_x")+
  geom_point(aes(position, logratio),
             data=selected.profiles,
             shape=1)+
  scale_x_continuous(
    "position on chromosome",
    breaks=c(100, 200))+
  scale_y_continuous(
    "logratio",
    limits=c(-1,1)*1.1)
print(gg.unsupervised)
```



\vspace{1cm}

Now, fitting the unsupervised change point model using the cpt.mean function and using 3 penalty paramters "SIC0, BIC0 and AIC0" we get:


```{r}
pen.name <- "SIC0"
(models <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ),
    before_mean=mean(logratio[1:before.change]),
    after_mean=mean(logratio[after.change:length(logratio)]),
    end_pos=position[length(position)])
}, by=list(profile.id, chromosome)])

pen.name <- "BIC0"
(models1 <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ),
    before_mean=mean(logratio[1:before.change]),
    after_mean=mean(logratio[after.change:length(logratio)]),
    end_pos=position[length(position)])
}, by=list(profile.id, chromosome)])

models <- rbind(models, models1)


pen.name <- "AIC0"
(models2 <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ),
    before_mean=mean(logratio[1:before.change]),
    after_mean=mean(logratio[after.change:length(logratio)]),
    end_pos=position[length(position)])
}, by=list(profile.id, chromosome)])

models <- rbind(models, models2)


changes <- models[, data.table(
  change=changes[[1]],before_mean = before_mean, 
  after_mean = after_mean, end_pos=end_pos,
  start = 0, end = changes[[1]], mean = before_mean
), by=list(profile.id, chromosome, pen.name)]

segments <- changes[, data.table(
  change=change,before_mean = before_mean, 
  after_mean = after_mean, end_pos=end_pos,
  start = change, end = end_pos, mean = after_mean
), by=list(profile.id, chromosome, pen.name)]

segments = rbind(changes, segments)
```

\vspace{1cm}
Plotting the data along with optimal segment means (green line segments) we get:


```{r}
gg.unsupervised+
  facet_grid(pen.name ~ profile.id + chromosome, scales="free", space="free_x")+
  theme(legend.box="horizontal")+
  geom_vline(aes(
    xintercept=change),
    color="green",
    size=1,
    linetype="dashed",
    data=segments)+
  geom_segment(aes(
    x = start,
    y = mean,
    xend = end,
    yend = mean),
    col = I("green"),
    data = segments
  )
```


**Test 3**


Performing test on all chromosomes for profile ids in [1 100].


```{r}
num_tests = 100
#selected_data = filter(neuroblastoma$profiles, chromosome == sel_chromosome)
selected_ids = unique(neuroblastoma$profiles$profile.id)[1:num_tests]
selected_data = filter(neuroblastoma$profiles, profile.id %in% selected_ids)
```


Performing the test:


```{r}
timing_list <- list()
n = 1
while(n <= num_tests){
  current_id = selected_ids[n]
  current_data = filter(selected_data, profile.id == current_id)
  length = length(current_data$logratio)
  timing <- microbenchmark(
    "cpt_mean"={
        cpt.mean(current_data$logratio, method="PELT", pen.value = 1)
    }, 
  "fpop"={
      Fpop(current_data$logratio, 1)
    }, times=5)
  
  timing_list[[paste(n)]] <- data.table(length, timing)
  n = n + 1
}

timing.dt <- do.call(rbind, timing_list)

ggplot(data = timing.dt, aes(x = log(timing.dt$length), y = log(timing.dt$time), col = timing.dt$expr))+
  geom_smooth()+
  labs(x="log(size)", y="log(time)", col="method")
```