---
title: "GSOC Tests"
author: "Anuraag Srivastava(as4378)"
date: "February 22, 2019"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

**Test 1**

Run either changepoint::cpt.mean or Fpop::fpop on one of the data sets (vector of logratio values for a given profile.id/chromosome combination) in neuroblatoma$profiles from data(neuroblastoma, package="neuroblastoma"). For one penalty parameter, plot the data as black points and the optimal segment means as horizontal green line segments.

\vspace{1cm}
Loading the required packages:


```{r}

data(neuroblastoma, package="neuroblastoma")
options(width=100)
```
\vspace{1cm}

Selecting one profile id to continue with the test. This profile is of a children status "relapse" several years after treatment, hence a good candidate for change point detection problem.


```{r}
selected <- data.frame(
  profile.id=paste(c(1)),
  status=c("relapse"))
selected
```
\vspace{1cm}

Creating a data table for selected profile:


```{r}
rownames(selected) <- selected$profile.id
selected$status.profile <- with(selected, paste(status, profile.id))
some.ids <- rownames(selected)
library(data.table)
someProfiles <- function(all.profiles){
  some <- subset(all.profiles, paste(profile.id) %in% some.ids)
  status.profile <- selected[paste(some$profile.id), "status.profile"]
  some$status.profile <- ifelse(
    is.na(status.profile), paste(some$profile.id), status.profile)
  data.table(some)
}
selected.profiles <- someProfiles(neuroblastoma$profiles)
```

\vspace{2cm}

Now, for selected profile there are 24 change-point detection problems (24 chromosomes). Plotting this problem in a grid we get:


```{r}
library(ggplot2)
gg.unsupervised <- ggplot()+
  ggtitle("unsupervised changepoint detection = only noisy data sequences")+
  theme(
    panel.margin=grid::unit(0, "lines"),
    panel.border=element_rect(fill=NA, color="grey50")
  )+
  facet_grid(status.profile ~ chromosome, scales="free", space="free_x")+
  geom_point(aes(position/1e6, logratio),
             data=selected.profiles,
             shape=1)+
  scale_x_continuous(
    "position on chromosome (mega bases)",
    breaks=c(100, 200))+
  scale_y_continuous(
    "logratio (approximate DNA copy number)",
    limits=c(-1,1)*1.1)
print(gg.unsupervised)
```


Now, fitting the unsupervised change point model using the cpt.mean function and using "SIC0" as the penalty parameter we get:

\vspace{1cm}
```{r}
pen.name <- "SIC0"
(unsupervised.models <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ))
}, by=list(profile.id, status.profile, chromosome)])


(unsupervised.changes <- unsupervised.models[, data.table(
  change=changes[[1]]
), by=list(profile.id, status.profile, chromosome, pen.name)])
```


\vspace{1cm}
Plotting the data along with optimal segment means (green line segments) we get:


```{r}
gg.unsupervised+
  theme(legend.box="horizontal")+
  geom_vline(aes(
    xintercept=change/1e6),
    color="green",
    size=1,
    linetype="dashed",
    data=unsupervised.changes)
```



**Test 2**

For two data sets and three penalty parameters, plot the data and optimal models using a ggplot with facet_grid(segments ~ profile.id + chromosome)


\vspace{1cm}

This time selecting 2 profiles with status "relapse".


```{r}
selected <- data.frame(
  profile.id=paste(c(1, 4)),
  status=c("relapse", "relapse"))
selected
```

\vspace{1cm}
Creating data table for selected profiles:


```{r}
rownames(selected) <- selected$profile.id
selected$status.profile <- with(selected, paste(status, profile.id))
some.ids <- rownames(selected)
library(data.table)
someProfiles <- function(all.profiles){
  some <- subset(all.profiles, paste(profile.id) %in% some.ids)
  status.profile <- selected[paste(some$profile.id), "status.profile"]
  some$status.profile <- ifelse(
    is.na(status.profile), paste(some$profile.id), status.profile)
  data.table(some)
}
selected.profiles <- someProfiles(neuroblastoma$profiles)
```

\vspace{1cm}
Printing the problems (24 * 2 = 48):

```{r}
library(ggplot2)
gg.unsupervised <- ggplot()+
  ggtitle("unsupervised changepoint detection = only noisy data sequences")+
  theme(
    panel.margin=grid::unit(0, "lines"),
    panel.border=element_rect(fill=NA, color="grey50")
  )+
  facet_grid(status.profile ~ chromosome, scales="free", space="free_x")+
  geom_point(aes(position/1e6, logratio),
             data=selected.profiles,
             shape=1)+
  scale_x_continuous(
    "position on chromosome (mega bases)",
    breaks=c(100, 200))+
  scale_y_continuous(
    "logratio (approximate DNA copy number)",
    limits=c(-1,1)*1.1)
print(gg.unsupervised)
```



\vspace{1cm}

Now, fitting the unsupervised change point model using the cpt.mean function and using 3 penalty paramters "SIC0, BIC0 and AIC0" we get:


```{r}
pen.name <- "SIC0"
(unsupervised.models <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ))
}, by=list(profile.id, status.profile, chromosome)])

pen.name <- "BIC0"
(unsupervised.models1 <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ))
}, by=list(profile.id, status.profile, chromosome)])

unsupervised.models <- rbind(unsupervised.models, unsupervised.models1)


pen.name <- "AIC0"
(unsupervised.models2 <- selected.profiles[, {
  fit.pelt <- changepoint::cpt.mean(
    logratio, penalty=pen.name, method="PELT")
  end <- fit.pelt@cpts
  before.change <- end[-length(end)]
  after.change <- before.change+1L
  data.table(
    pen.name,
    pen.value=fit.pelt@pen.value,
    changes=list(
    as.integer((position[before.change]+position[after.change])/2)
    ))
}, by=list(profile.id, status.profile, chromosome)])

unsupervised.models <- rbind(unsupervised.models, unsupervised.models2)


(unsupervised.changes <- unsupervised.models[, data.table(
  change=changes[[1]]
), by=list(profile.id, status.profile, chromosome, pen.name)])
```

\vspace{1cm}
Plotting the data along with optimal segment means (green line segments) we get:


```{r}
gg.unsupervised+
  facet_grid(pen.name ~ profile.id + chromosome, scales="free", space="free_x")+
  theme(legend.box="horizontal")+
  geom_vline(aes(
    xintercept=change/1e6),
    color="green",
    size=1,
    linetype="dashed",
    data=unsupervised.changes)

```

